---
title: "A new approach to extrapolation: blended survival curves  "
author: "Zhaojing Che | Department of Statistical Science | University College London"
#         | [zhaojing.che.19@ucl.ac.uk](mailto:zhaojing.che.19@ucl.ac.uk)
# date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
  bookdown::word_document2:
  bookdown::pdf_book:
    keep_tex: yes
#    reference_docx: BMJ.docx
  pdf_document: default
header-includes: \usepackage{bm}
fig_caption: yes
link-citations: yes
bibliography: references.bib 
---

# Motivating case study  {-}

Here is an example about the technical implementation (including relevant `R` code) of the blended model. Firstly, the “blending” idea is to consider two separate processes to describe the long-term
horizon survival; one is to use a flexible model (e.g. Cox semi-parametric) to fit as well as possible the observed data, and the other is a parametric model encoding assumptions on the expected behaviour of underlying long-term survival. Then importantly, the two are “blended” into a single survival curve that is identical with the flexible model over the range of observed times and increasingly similar to the parametric model over the extrapolation period.  

## Install and load required R packages {-}

We will use two R packages:

- `INLA`    for computing observed survival estimates based on internal data 
- `survHE`  for computing external survival estimates with expert judgement

Install the packages
```{r install packages, eval=TRUE}
if (.Platform$OS.type == "windows") {
  if ("survHE" %in% rownames(installed.packages()) == FALSE) {
    install.packages("survHE",
                     repos = c("http://www.statistica.it/gianluca/R",
                               "https://cran.rstudio.org",
                               "https://inla.r-inla-download.org/R/stable"),
 	             dependencies = TRUE)
   }
}

if (.Platform$OS.type == "windows") {
  if ("INLA" %in% rownames(installed.packages()) == FALSE) {
    install.packages("INLA",
                     repos = c(getOption("repos"),
                     INLA = "https://inla.r-inla-download.org/R/stable"),
                     dep = TRUE)
   }
}
```

Load the packages
```{r load packages, results='hide', warning=FALSE, message=FALSE}
library(survHE)
library(INLA)
```

## Example dataset (TA174)    {-}
We will use CLL-8 trial data in NICE technology appraisal TA174, which compares rituximab in combination with fludarabine and cyclophosphamide
(FCR) to fludarabine and cyclophosphamide (FC) for the first-line treatment of chronic lymphocytic leukemia. Here we make the arm FCR as the example.
 
```{r load data}
load("../Data/TA174_FCR.RData")
head(dat_FCR)
```
- `patid`: Patient ID
- `treat`: Treatment arm 1 = FCR
- `death`: Death status 0 = censored, 1 = dead
- `death_t`: Survival time in months
- `death_ty`: Survival time in years

Before starting the analysis, we need to firstly source a R script that contains all functions required in the following components.   
```{r source function, include=TRUE}
source("../Scripts/Functions.R", local = knitr::knit_global())
```

##  Observed time period   {-}

### Internal curve: piecewise exponential model  {-}

To provide the best fit possible and a good level of flexibility, we will use a Cox semi-parametric model with piecewise constant hazard. The hazard is modeled as 
$$ h(t) = \text{exp}(\lambda_{k}); \quad t\in (u_{k-1}, u_k], \quad k = 1, \ldots, K $$
Hence the time axis is partitioned into $K$ intervals using $0<u_1<u_2<\ldots<u_K<\infty$ and variables $\mathbf{\lambda} = (\lambda_1, \ldots, \lambda_K)$ are assumed to a prior with random walks (RW) of order one (or two). 

The `surv_est_inla` function creates internal survival curves ($S_{obs}$) based on the model fitted to the observed data using Bayesian INLA approach. Below, we set the time horizon from 0 to 180 months (15 years), during which intervals are constructed for every five months in the option `cutpoints`.   

```{r observed estimate}
obs_Surv <- surv_est_inla(data = dat_FCR, 
                          cutpoints = seq(0,180, by =5))
                          
names(obs_Surv)
```

Two key components of `obs_Surv` object that will be used to plot survival curves:

- `time`:  times used to create the survival curves.
- `S_obs`: survival probabilities corresponding to elements of `time`

Figure \@ref(fig:observedplot) shows that the model achieves a
feasible fit to the observed data, but does not generate a credible extrapolation. It suggests over 50% survival at 15 years (end of the green curve), which is in stark contrast with
expert estimates, suggesting instead that only 1.3% of the cohort would be likely to survive beyond that
time. 


```{r observedplot, echo=TRUE, fig.cap="Survival estimates of the FCR arm based on the piecewise exponential model fitted to the observed data"}
ggplot() + 
  geom_line(aes(obs_Surv$KM$time, obs_Surv$KM$surv, colour = "Kaplan-Meier"), size = 1.25, linetype = "dashed") +
  geom_line(aes(obs_Surv$time, rowMeans(obs_Surv$S_obs), colour = "Data fitting"), size = 1) +
  geom_ribbon(aes(x=obs_Surv$time, y = rowMeans(obs_Surv$S_obs), 
                  ymin = apply(obs_Surv$S_obs, 1, quantile, probs = 0.025),
                  ymax = apply(obs_Surv$S_obs, 1, quantile, probs = 0.975)), alpha = 0.1) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), axis.line = element_line(colour = "black"), 
        text = element_text(size = 8)) +
  xlab("Time (months)") + ylab("Survival") +
  xlim(0, 180) + ylim(0,1) +
  scale_colour_manual(name = "model", values = c("Data fitting"="#7CAE00", "Kaplan-Meier"="#990000"))
```

##  Unobserved time period   {-}

### External curve: using expert judgement  {-}

To produce a reasonable and realistic estimate for the long-term survival, we consider an external curve that is not informed by the observed data, but driven by the evidence from a different data source instead. Ideally, we can access matched long-term data from a relevant study, possibly of an observational nature, such as a registry or a cohort study, yet it is not always the case. Here we focus on a more general situation when only subjective knowledge is available, typically in the form of expert elicitation.    

The `ext_surv_est` function creates external survival curves ($S_{ext}$) encoding external assumptions from expert opinion. In this motivating example, there is a strong opinion that approximately 1.3% of the cohort would be alive beyond 15 years, hence we identify the lifetime horizon as 15 years (180 months) [@roche2008rituximab]. Then if it is assumed that the survival probability at 12 years (`t_info = 144`) should be approximately 5% (`S_info = 0.05`) based on previous experience, we can use following code:

```{r external estimate, message=FALSE, warning=FALSE, results='hide'}
ext_Surv <- ext_surv_est(t_info = 144, S_info = 0.05, 
                         T_max = 180, 
                         times_est = seq(0, 180), distr = "gom")

names(ext_Surv)
```

```{r external_object, echo=FALSE}
names(ext_Surv)
```

## Blended estimate      {-}

Given both internal and external curves over the whole time-horizon, we can compute the blended estimate based on a weight function $\pi(t; \alpha, \beta, a, b)$. 

Here we set the blending interval $[48, 150]$ and $Beta(3, 3)$. The object `weight` contains the values of weight allocated to the external survival over the time.  
```{r weight function}
blending_interval <- list(min = 48, max = 150)
beta_params <- list(alpha = 3, beta = 3)

tp <- seq(0, 180)

# parameters for the weight function
wt_par <- list(a = blending_interval$min, b = blending_interval$max,
               shape1 = beta_params$alpha, shape2 = beta_params$beta) 

weight <- with(wt_par,
               pbeta((tp - a)/(b - a), shape1, shape2))
```

The blended survival curve is simply obtained as
$$S_{ble}(t) = S_{obs}(t)^{1-\pi(t; \alpha, \beta, a, b)}\times S_{ext}(t)^{\pi(t;\alpha, \beta, a, b)} $$
```{r blended estimate}
n_sim <- ncol(ext_Surv$S_ext)

ble_Surv <- matrix(NA, nrow = 180 + 1, ncol = n_sim)

for (i in seq_len(n_sim)) {
  ble_Surv[, i] <-
    obs_Surv$S_obs[, i]^(1 - weight) * ext_Surv$S_ext[, i]^weight
}

```

Figure \@ref(fig:blendedplot) shows the blended survival estimate of the FCR arm driven by internal and external curve over the whole time-horizon. Below represents the result of one assumption about the blending operation, and different scenarios can be tested by changing the values of the parameters (`wt_par`) associated with the weight function.

```{r blendedplot, echo=TRUE, fig.cap="Blended survival curve based on short-term data and external information for FCR arm"}
ggplot() + 
  geom_line(aes(obs_Surv$KM$time, obs_Surv$KM$surv, colour = "Kaplan-Meier"), size = 1.25, linetype = "dashed") +
  geom_line(aes(tp, rowMeans(obs_Surv$S_obs), colour = "Data fitting"), 
            size = 1, linetype = "twodash") +
  geom_ribbon(aes(x=tp, y = rowMeans(obs_Surv$S_obs),
                  ymin = apply(obs_Surv$S_obs, 1, quantile, probs = 0.025),
                  ymax = apply(obs_Surv$S_obs, 1, quantile, probs = 0.975)), alpha = 0.1) +
  geom_line(aes(tp, rowMeans(ext_Surv$S_ext), colour = "External info"), 
            size = 1, linetype = "longdash") +
  geom_ribbon(aes(x=tp, y = rowMeans(ext_Surv$S_ext), 
                  ymin = apply(ext_Surv$S_ext, 1, quantile, probs = 0.025),
                  ymax = apply(ext_Surv$S_ext, 1, quantile, probs = 0.975)), alpha = 0.1) +
  geom_line(aes(tp, rowMeans(ble_Surv), colour = "Blended curve"), size = 1.25) +
  geom_ribbon(aes(x=tp, y = rowMeans(ble_Surv), 
                  ymin = apply(ble_Surv, 1, quantile, probs = 0.025),
                  ymax = apply(ble_Surv, 1, quantile, probs = 0.975)), alpha = 0.1) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(), axis.line = element_line(colour = "black"), 
        text = element_text(size = 8)) +
  xlab("Time (months)") + ylab("Survival") +
  scale_colour_manual(name = "model", values = c("Data fitting"="#7CAE00", "External info"="#00BFC4", "Blended curve"="#F8766D", "Kaplan-Meier"="#990000"))
```

# References {-}
